---
title: "Accessing LEHD data"
date:  "2022-11-28"
author: "Coach Skufca"
output: html_notebook
---

This notebooks will try to use the `lehdr` package to get at detailed demographic data (where are interest is understanding how many people WORK in each census tract in the DC area.)

Primary resource is https://cran.r-project.org/web/packages/lehdr/vignettes/getting_started.html

## Install lehdr


Run this code only once.


```{r}
#devtools::install_github("jamgreen/lehdr")
```


## Packages

Standards:

```{r}
library(knitr)
library(tidyverse)
library(janitor)
library(lubridate) # because we will probably see some dates
library(here) # a package I haven't taught you about before that doesn't do much, but ....
```

Some additional packages focused on today's work:

```{r}
library(sf) # working with simple features - geospatial
library(tmap)
library(tidycensus)
library(lehdr)

```
## Informational resources for mapping

* An overall resource on mapping in R: https://bookdown.org/nicohahn/making_maps_with_r5/docs/introduction.html
* A starting point to learn about `sf`:  https://r-spatial.github.io/sf/articles/
* Getting started with `tmap`: https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html
* The `tidycensus` package: https://walker-data.com/tidycensus/index.html
* The book on `tidycensus` : https://walker-data.com/census-r/index.html


# From Vignette

We will apply some of the ideas from the vignette: 

https://cran.r-project.org/web/packages/lehdr/vignettes/getting_started.html

One of the additional resources specified is this document which describes all of the variables in the data:

https://lehd.ces.census.gov/data/lodes/LODES7/LODESTechDoc7.5.pdf


## Proof of concept

*Taking material and code directly from the vignette:*


This first example pulls the Oregon (state = "or") 2014 (year = 2014), origin-destination (lodes_type = "od"), all jobs including private primary, secondary, and Federal (job_type = "JT01"), all jobs across ages, earnings, and industry (segment = "S000"), aggregated at the Census Tract level rather than the default Census Block (agg_geo = "tract").


```{r}
or_od <- grab_lodes(state = "or", year = 2014, lodes_type = "od", job_type = "JT00", 
           segment = "S000", state_part = "main", agg_geo = "tract")

head(or_od)
```

The package can be used to retrieve multiple states and years at the same time by creating a vector or list. This second example pulls the Oregon AND Rhode Island (state = c("or", "ri")) for 2013 and 2014 (year = c(2013, 2014) or year = 2013:2014).


```{r}
or_ri_od <- grab_lodes(state = c("or", "ri"), year = c(2013, 2014), lodes_type = "od", job_type = "JT01", 
           segment = "S000", state_part = "main", agg_geo = "tract") 
```

    

head(or_ri_od)
Not all years are available for each state. To see all options for lodes_type, job_type, and segment and the availability for each state/year, please see the most recent LEHD Technical Document at https://lehd.ces.census.gov/data/lodes/LODES7/.

Other common uses might include retrieving Residential or Work Area Characteristics (lodes_type = "rac" or lodes_type = "wac" respectively), low income jobs (segment = "SE01") or good producing jobs (segment = "SI01"). Other common geographies might include retrieving data at the Census Block level (agg_geo = "block", not necessary as it is default) – but see below for other aggregation levels.


## Some DC data:

Let's get lodes for DC.

We will use `state_part="aux"` so that we get people working in DC who don't live there.

We look at JT01 (primary job - the job that provides each person with the most earnings.)


```{r}
dc_od <- grab_lodes(state ="dc", year = 2019, lodes_type = "od", job_type = "JT01", 
           segment = "S000", state_part = "aux", agg_geo = "tract",use_cache=TRUE)

head(dc_od)
```
Total workers:

```{r}
sum(dc_od$S000)

```




Let's pull only workplace associate dataset:

* "wac" aggegates jobs based on job location
* "rac" aggreates worker based on their home location.

```{r}
dc_wac <- grab_lodes(state = c("dc"), year = 2019, lodes_type = "wac", job_type = "JT00", 
           segment = "S000", state_part = "aux", agg_geo = "tract") %>% filter(state=="DC")

dc_rac <- grab_lodes(state = c("dc"), year = 2019, lodes_type = "rac", job_type = "JT00", 
           segment = "S000", state_part = "aux", agg_geo = "tract") %>% filter(state=="DC")

head(dc_wac)
```

How many people work in DC?

```{r}
sum(dc_wac$C000)
sum(dc_rac$C000)
```

Note the difference between the two sums --- # or workers vs # of workers who live in that area.

## Get some data from tidycensus using census tracts


Load the census key and get variables:
```{r}
census_api_key("4c7e5b613afdf7d2fd6d80005f53282c9a7c1bae")

#what variables
v20 = load_variables(2019,"acs5",cache=TRUE)
# median_family_income="	B06011_001" 
# all "B00001_001"	
#black "B02009_001"
```

```{r}
df_census=get_acs(geography = "tract",
                  variables=c(
                              "total_pop"="B01003_001",
                              "worker_living"="B08301_001"),
                  state="DC",geometry=TRUE,year=2019) 
```


```{r}
df_cens=df_census %>% rename(val=estimate) %>%  
  pivot_wider(names_from = "variable",
              values_from = c("val","moe")) %>%
  clean_names()


tmap_mode("view")
tm_shape(df_cens) +tm_polygons("val_total_pop",alpha=.5)
```

### An interesting anomoly

Note that the two datasets give different answers for number of workers living in the area.

Partly 
```{r}
sum(df_cens$val_worker_living)
sum(dc_rac$C000)
```

This difference may be due to differences in accounting (as some people are not counted in the LODES data).

But ... we will ignore that discrency for now.


### Merge census and lodes data

By joining wac data to census data, we now have a dataframe will all the data AND the geometry by tract.

```{r}
df_c=df_cens %>% left_join(dc_wac,by=c("geoid"="w_tract"))
```


##  Some viz

I build here some vizualizations that I found interesting.  The first looks at where men work vs where women work.

```{r}
df_c1=df_c %>% select(pop=val_total_pop, men=CS01,women=CS02) %>% #renames while selecting
  mutate(ratio=women/men)


df_c1%>%
  tm_shape()+tm_polygons(c("pop","men","women"),alpha=.4)




```

And below, we simply examine the ratio between those groups.

```{r}
df_c1%>%
  tm_shape()+tm_polygons(c("ratio"),alpha=.4)
```

A third vizualization looks at racial/ethnic differences where workers work.

```{r}
df_c %>% select(white_workers=CR01, black_workers=CR02,hispanic_workers=CT02) %>%
  tm_shape()+tm_polygons(c("white_workers","black_workers","hispanic_workers"),alpha=.4)
```









