---
title: "Joining spatial data to study bikeshare."
date:  "2022-11-09"
author: "Coach Skufca"
output: html_notebook
---

This notebooks will step through the two tasks issued in class on Monday, 14 Nov.

* Get data using tidycensus so that we can understand "where are people during the day?"
* Visualize bikeshare activity by DC health neighborhood.

## Packages

Standards:

```{r}
library(knitr)
library(tidyverse)
library(janitor)
library(lubridate) # because we will probably see some dates
library(here) # a package I haven't taught you about before that doesn't do much, but ....
```

Some additional packages focused on today's work:

```{r}
library(sf) # working with simple features - geospatial
library(tmap)
library(tidycensus)

```
## Informational resources

* An overall resource on mapping in R: https://bookdown.org/nicohahn/making_maps_with_r5/docs/introduction.html
* A starting point to learn about `sf`:  https://r-spatial.github.io/sf/articles/
* Getting started with `tmap`: https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html
* The `tidycensus` package: https://walker-data.com/tidycensus/index.html
* The book on `tidycensus` : https://walker-data.com/census-r/index.html


## Using the Neighborhood Geospatial Data (using /data)


Our first data source comes from opendata.dc

https://opendata.dc.gov/datasets/DCGIS::dc-health-planning-neighborhoods/about


I will use the GeoJSON file.  (Newer, not necessarily better, but ... a single file.  Not smaller, but ... this one is not big.)  




Data is easily readable 
```{r}
neigh=st_read(here("data_raw","dc_neigh.geojson")) %>% clean_names() %>%
   mutate(shapearea=st_area(neigh))
class(neigh)


```

```{r}
plot(neigh)
```



## Reminder - Joins



## Investigating joining spatial and non-spatial data

Read basic Sep2022 data.

```{r}
rdsfile=here("data","sep2202_bikes_cleaned.rds")

dfb_rds=readRDS(rdsfile)
```

Filter to rides on 1 Sep, in 1st 4 hours, to give a small dataset. 
```{r}
#dfb1=dfb_rds %>% filter(mday(start)==1,hour_of_day<=4)
dfb1=dfb_rds %>% filter(mday(start)==1)
```



## Spatial joining (of dataframes)

Create a sf object from the bikedata

```{r}
dfb_sf=st_as_sf(dfb1, coords = c("start_lng", "start_lat"), 
                 crs = 4326)
```

Join datasets


```{r}
neigh_temp=st_join(dfb_sf,neigh,join = st_within) %>% group_by(code) %>% summarise(n=n()) %>%
  as_tibble() %>% select(-geometry)

neigh2=neigh %>% left_join(neigh_temp)

```



```{r}

tmap_mode("view")

tm_shape(neigh2) +tm_polygons("n",alpha=.5,n=6)
```


## Joining with other spatial data

Let's get some data using `tidycensus`.  Need an API key   https://api.census.gov/data/key_signup.html


```{r}


census_api_key("4c7e5b613afdf7d2fd6d80005f53282c9a7c1bae")

#what variables
v20 = load_variables(2018,"acs5")
# median_family_income="	B06011_001" 
# all "B00001_001"	
#black "B02009_001"
```


Get some data:

```{r}
df_cencus=get_acs(geography = "tract",
                  variables=c("median_inc"="B06011_001",
                              "pop"="B01001_001",
                              "pop_black"="B02009_001"),
                  state="DC",geometry=TRUE,year=2018) 
```

```{r}
class(df_cencus)
plot(df_cencus)
```
It's in long format.  Let's make it wide.
```{r}
df_cens=df_cencus %>% select(-moe) %>% spread(variable,estimate) 

tm_shape(df_cens) +tm_polygons("median_inc",alpha=.5)
```


```{r}

  tm_shape(neigh2) +tm_borders(col="blue",lwd=5,alpha=.2)+
  tm_shape(df_cens) +tm_borders(col="red",lwd=1,alpha=.3)
```



```{r}
#<<<<<<< HEAD
#df_j=st_join(df_cens,neigh2)
#=======
#df_j=st_join(df_cens,neigh2,prepared=FALSE)
#>>>>>>> aaf01be5cf721819dd2df615aef7a1999bcec0c2
```

```{r}
df_cens_adj=df_cens %>% st_transform(4326)
```

```{r}
df_j=st_join(df_cens_adj,neigh2,largest=TRUE)
```
Other order?:

```{r}
#<<<<<<< HEAD
##df_j_rev = st_join(neigh2,df_cens_adj,largest=TRUE)
#=======
#df_j_rev = st_join(neigh2,df_cens_adj,largest=TRUE)
#>>>>>>> aaf01be5cf721819dd2df615aef7a1999bcec0c2
```

Since we want the geometry for the NEIGHBORHOODS, we need a to work a little harder:

```{r}
df1=df_j %>% select(median_inc,pop,pop_black,objectid) %>%
  group_by(objectid) %>%
  summarise(pop_n=sum(pop),
            pop_black_n=sum(pop_black), 
            adj_median_income=sum(pop*median_inc)/pop_n) 

plot(df1)
```

```{r}
#df2=left_join(neigh2,df1)

df2=left_join(neigh2,df1 %>% st_set_geometry(NULL))

```

```{r}
df2=df2 %>% mutate(black_perc=pop_black_n/pop_n, covid_rate=total_positives/pop_n)
tm_shape(df2)+tm_polygons(c("adj_median_income","covid_rate","black_perc"))
```



```{r}
df2 %>% filter(objectid!=30) %>% tm_shape()+tm_polygons(c("adj_median_income","covid_rate","black_perc"),alpha=.4)
```


