---
title: "NYC Flights"
author: "Coach Skufca"
date: "2022-09-09"
output: html_notebook
---

Class play with NYC flights data.

## Load Packages

```{r}
library(nycflights13)
library(tidyverse)
library(here)
library(janitor)
```

## Load flight data

```{r}
df1=flights


```

## Class work

```{r}
df2=df1 %>% 
  filter(origin=="MIA",month==9)
```


```{r}
df3=df1 %>% 
  filter(dest=="MIA",month==9)
```

```{r}
df4=df1 %>% 
  filter(dest=="MIA",month==1)
```

```{r}
df5=df1 %>% 
  filter(dest=="MIA",month==7)
```


```{r}
df6=df1 %>% 
  filter(dest=="MIA")


df6 %>% ggplot(aes(month)) + geom_bar()
```
Analyzing the graph above we note:

* December has the most flights, perhaps due to holiday.
* Feb is low, perhaps due to fewer days in Feb.
* Flight numbers seem to correlate with days in the month.













```{r}
df7=df1 %>% 
  filter(dest %in% c("LAX","SAN","SFO","ATL","ORD","MIA"))


df7 %>% ggplot(aes(x=month,fill=dest)) + geom_bar()
```


```{r}

df7 %>% ggplot(aes(month,fill=dest)) + geom_bar()+facet_wrap(~dest)
```
Observation from above graph:

* San Diego is much smaller than the others.
* MIA is _unusually_ flat.
* ORD is summer intensive (more traffic over summer)
* ORD, LAX, SFO patterns similar: low in winter, high in summer.
* ATL is flat, perhaps it's a hub.
* Feb is low (fewer days) and winter.







```{r}
thisfile=here("data_raw","DL_SelectFields.zip")

df2019=read_csv(thisfile) %>% clean_names()
```

```{r}
df9=df2019 %>% filter(passengers>0,origin=="LGA")

df9 %>% ggplot(aes(month)) + geom_bar()
```

```{r}
df9=df2019 %>% filter(passengers>0,origin=="LGA")


df9 %>% ggplot(aes(month)) + geom_bar(aes(weight=departures_performed))
```

```{r}
df9=df2019 %>% filter(passengers>0,origin=="LGA"|dest=="LGA")

df9 %>% ggplot(aes(month)) + geom_bar(aes(weight=passengers))
```


```{r}
df9=df2019 %>% filter(passengers>0,origin=="LGA")

df9 %>% ggplot(aes(month)) + geom_bar(aes(weight=departures_performed))
```
